A point-level trend model for the North American Breeding Bird Survey that corrects for detectability

### 2024-08-30 article/article.tex
completed second draft of chapter 4

\par All biological data, including long-term biological data, have challenges. 
One of the principal challenges in any biological data collection is detectability \citep{bennett_how_2024}.
In the BBS, unmodeled or uncontrolled changing of detectability of birds through time can bias estimates of trends from BBS models.
The BBS field methods currently control for many factors that are known to affect detectability, such as observer effects \citet{sauer_observer_1994}.
Some sources of potential changing detectability can be modeled using route-level summaries, including changes in traffic during surveys \citep{griffith_traffic_2010} and changes in phenology (CITE ENGLISH).
However many of the factors affecting detectability may vary among the points within a given route.
For example, changes in forest cover through time at a specific point within a route will result in changes in detectability at that point.
If forest coverage increases, then detectability for many bird species will generally decrease, and vice versa if forest coverage decreases \citep{edwards_point_2023, yip_sound_2017}.

@@

\par The following subsections describe how the current route-level description of the model can be converted to a point-level model, by taking advantage of the point-level data provided by the United States Geological Survey and Canadian Wildlife Service that has been available since 1997. 
We first describe the current status and trend model used by the Canadian Wildlife Service, as well as relevant Bayesian priors and assumptions with that model.
We then describe how to use the point-level data in this status and trend model, as well as describe additional choices a researcher could make regarding route-level hyperparameters.
Then, we expand this point-level model to explicitly include detectability estimates, as well as uncertainty around those estimates using variance propagation techniques described by \citep{bravington_variance_2021}.
Table \ref{table:params} provides a reference guide to all the parameters used in these models.

\begin{table}[ht]\label{table:params}
	\centering
	\caption{Parameters in the BBS models.}
		\begin{tabular}[t]{l>{\raggedright}p{0.5\linewidth}>{\raggedright\arraybackslash}p{0.3\linewidth}}
		\hline
		Parameter or Datum&Description&Model\\
		\hline
		$i$&Stratum index&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$j$&Point/stop index&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$J$&Route index&Equation \ref{bbs}\\
		$t$&Year index&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$o$&Observer index&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$Y_{J,i,t,o}$&Total observed birds at route $J$, stratum $i$, year $t$, by observer $o$&Equation \ref{bbs}\\
		$Y_{j,i,t,o}$&Total observed birds at point $j$, stratum $i$, year $t$, by observer $o$&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\alpha_i$&Stratum-level intercept, modelled with an intrinsic conditional autoregressive model \citep{besag_bayesian_1991, smith_spatially_2023}&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\delta_{i,t}$&Smoothed temporal component estimating population trajectory through time. Modelled with generalized additive model \citep{smith_north_2021}&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\gamma_{i,t}$&Annual year-effect fluctuation term; departure from smoothed GAM&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\psi_J$&Route-level intercept&Equation \ref{bbs}\\
		$\psi_j$&Point-level intercept&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\omega_o$&Observer-level intercept&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\eta$&First-year observer effect&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$I(o,J,t)$&1 if observer is surveying route $J$ in year $t$ for the first time, 0 otherwise&Equation \ref{bbs}\\
		$I(o,j,t)$&1 if observer is surveying point $j$ in year $t$ for the first time, 0 otherwise&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$p_{j,i,t}$&Probability of availability at point $j$ in stratum $i$ in year $t$&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$q_{j,i,t}$&Probability of perceptibility at point $j$ in stratum $i$ in year $t$&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\kappa_{j,i,t}^{(p)}$&Design matrix for availability variance propagation; first derivatives of probability of availability \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
		$\kappa_{j,i,t}^{(q)}$&Design matrix for perceptibility variance propagation; first derivatives of probability of perceptibility \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
		$\zeta$&Effects vector for availability variance propagation \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
		$\xi$&Effects vector for perceptibility variance propagation \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
		\hline
	\end{tabular}
\end{table}%

\subsection{Current BBS Protocol and Analysis}

@@

\subsection{Using Point-level Data in the BBS Status and Trend Models}

@@

\subsection{A Point-level Model that Explicitly Includes Detectability}

@@

\begin{figure}[h]
	\includegraphics{abundance-detectability_scaled.png}
	\caption{TO DO}
	\label{fig:example}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../output/plots/route-vs-point.png}
	\caption{Index of abundance (A) and trend (B) derived from POINT model versus ROUTE model. Each point in A represents an index of abundance for a stratum x year combination, whereas each point in B represents a trend estimate for a stratum. Regression lines (black) and posterior draws (grey) indicate modelled relationship.}
	\label{fig:route-vs-point}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../output/plots/precision_comparison.png}
	\caption{TO DO}
	\label{fig:precision}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../output/plots/varprop-diagnostics.png}
	\caption{Diagnostics of variance propagation as suggested by \citet{bravington_variance_2021}. A) Index of abundance as derived from the DETECT model versus index of abundance as derived from the VARPROP model, for each stratum x year combination (points). Regression line (black) and posterior draws (grey) indicate modelled relationship. B) Variance propagation parameter posterior distributions for availability. Because we are using the ``best" NA-POPS removal model for Ovenbird, we only have parameter distributions for the Intercept, Ordinal Day, and Ordinal Day squared terms. C) Variance propagation parameter posterior distributions for perceptibility. Similar to B), because we are using the ``best" NA-POPS distance model for Ovenbird, we include distributions for the Intercept, Forest, Road, and interaction terms.}
	\label{fig:varprop-diagnostics}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../output/plots/OVENpoint-vs-detectability-map.png}
	\caption{TO DO}
	\label{fig:route-change}
\end{figure}

>>>>

\par All biological data, including long-term biological data, have challenges.
One of the principal challenges in any biological data collection is detectability \citep{bennett_how_2024}.
In the BBS, unmodeled or uncontrolled changing of detectability of birds through time can bias estimates of trends from BBS models.
The BBS field methods currently control for many factors that are known to affect detectability, such as observer effects \citet{sauer_observer_1994}.
Some sources of potential changing detectability can be modeled using route-level summaries \cite{smith_patterns_2023}, including changes in traffic during surveys \citep{griffith_traffic_2010} and changes in phenology \citep{english_current_2021}.
However, many of the factors affecting detectability may vary among the points within a given route.
For example, changes in forest cover through time at a specific point within a route will result in changes in detectability at that point.
If forest coverage increases, then detectability for many bird species will generally decrease, and vice versa if forest coverage decreases \citep{edwards_point_2023, yip_sound_2017}.

@@

\par The following subsections describe how the current route-level description of the BBS status and trend model can be converted to a point-level model, by taking advantage of the point-level data provided by the United States Geological Survey and Canadian Wildlife Service that has been available since 1997. 
We first describe the current status and trend model used by the Canadian Wildlife Service, as well as relevant Bayesian priors and assumptions with that model.
We then describe how to use the point-level data in this status and trend model, as well as describe additional choices a researcher could make regarding route-level hyperparameters.
Then, we expand this point-level model to explicitly include detectability estimates, as well as uncertainty around those estimates using variance propagation techniques described by \citep{bravington_variance_2021}.
Table \ref{params_table} provides a reference guide to all the parameters used in these models.

\begin{table}[!ht]
	\caption{Parameters in the BBS models. Equation \ref{bbs} refers to the route-level model; Equation \ref{bbs-point} refers to the point-level model; Equation \ref{bbs-detect} refers to the point-level model with detectability offsets; and Equation \ref{bbs_varprop} refers to the point-level model with detectability and variance propagation.}
	\label{params_table}
		\begin{tabular}[t]{l>{\raggedright}p{0.5\linewidth}>{\raggedright\arraybackslash}p{0.3\linewidth}}
		\hline
		Parameter or Datum&Description&Model\\
		\hline
		$i$&Stratum index&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$j$&Point/stop index&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$J$&Route index&Equation \ref{bbs}\\
		$t$&Year index&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$o$&Observer index&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$Y_{J,i,t,o}$&Total observed birds at route $J$, stratum $i$, year $t$, by observer $o$&Equation \ref{bbs}\\
		$Y_{j,i,t,o}$&Total observed birds at point $j$, stratum $i$, year $t$, by observer $o$&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\alpha_i$&Stratum-level intercept, modelled with an intrinsic conditional autoregressive model \citep{besag_bayesian_1991, smith_spatially_2023}&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\delta_{i,t}$&Smoothed temporal component estimating population trajectory through time. Modelled with generalized additive model \citep{smith_north_2021}&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\gamma_{i,t}$&Annual year-effect fluctuation term; departure from smoothed GAM&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\psi_J$&Route-level intercept&Equation \ref{bbs}\\
		$\psi_j$&Point-level intercept&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\omega_o$&Observer-level intercept&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\eta$&First-year observer effect&Equation \ref{bbs}, \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$I(o,J,t)$&1 if observer is surveying route $J$ in year $t$ for the first time, 0 otherwise&Equation \ref{bbs}\\
		$I(o,j,t)$&1 if observer is surveying point $j$ in year $t$ for the first time, 0 otherwise&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$p_{j,i,t}$&Probability of availability at point $j$ in stratum $i$ in year $t$&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$q_{j,i,t}$&Probability of perceptibility at point $j$ in stratum $i$ in year $t$&Equation \ref{bbs-point}, \ref{bbs-detect}, \ref{bbs_varprop}\\
		$\kappa_{j,i,t}^{(p)}$&Design matrix for availability variance propagation; first derivatives of probability of availability \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
		$\kappa_{j,i,t}^{(q)}$&Design matrix for perceptibility variance propagation; first derivatives of probability of perceptibility \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
		$\zeta$&Effects vector for availability variance propagation \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
		$\xi$&Effects vector for perceptibility variance propagation \citep{bravington_variance_2021}&Equation \ref{bbs_varprop}\\
		\hline
	\end{tabular}
\end{table}%

\subsection{Current BBS Protocol and Analysis}\label{current-bbs}

@@

\subsection{Using Point-level Data in the BBS Status and Trend Models}\label{point-description}

@@

\subsection{A Point-level Model that Explicitly Includes Detectability}\label{detect-text}

@@

\begin{figure}[h]
	\includegraphics{abundance-detectability_scaled.png}
	\caption{Toy example of how changing forest coverage can affect both true abundance and observed abundance of a particular forest species. In Year 1, with high forest coverage, there may be ten individuals at a patch but only four may be detectable due to the high forest coverage. Ten years later, because of forest loss, there may only be four individuals, and three of them can be detected. This may appear as only a slight population decrease, when in reality the population decreased from ten to four at this site.}
	\label{fig:example}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../output/plots/route-vs-point.png}
	\caption{Index of abundance (A) and trend (B) derived from POINT model versus ROUTE model. Each point in A represents an index of abundance at the midyear in each stratum, whereas each point in B represents a trend estimate for a stratum. Regression lines (black) and posterior draws (grey) indicate modelled relationship.}
	\label{fig:route-vs-point}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../output/plots/precision_comparison.png}
	\caption{Boxplot of 90\% credible interval widths for trend estimates from each model.}
	\label{fig:precision}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../output/plots/varprop-diagnostics.png}
	\caption{Diagnostics of variance propagation as suggested by \citet{bravington_variance_2021}. A) Index of abundance as derived from the DETECT model versus index of abundance as derived from the VARPROP model, from the midyear from each stratum (points). Regression line (black) and posterior draws (grey) indicate modelled relationship. B) Variance propagation parameter posterior distributions for availability. Because we are using the ``best" NA-POPS removal model for Ovenbird, we only have parameter distributions for the Intercept, Ordinal Day, and Ordinal Day squared terms. C) Variance propagation parameter posterior distributions for perceptibility. Similar to B), because we are using the ``best" NA-POPS distance model for Ovenbird, we include distributions for the Intercept, Forest, Road, and interaction terms.}
	\label{fig:varprop-diagnostics}
\end{figure}

@@

\begin{figure}[h]
	\includegraphics{../output/plots/OVENpoint-vs-detectability-map.png}
	\caption{Difference in trend between the VARPROP and POINT model (i.e., VARPROP - POINT; A) and the DETECT and POINT model (i.e., DETECT - POINT; B).}
	\label{fig:route-change}
\end{figure}
